Crunching Large Graphs with Commodity Processors

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
programming model

Our shared-memory programming model partitions the overall address
space into a global shared address space and per node private address
spaces.  Locality may be exploited directly by the programmer in the
node private address spaces, just as in a conventional cache-coherent
multiprocessor node.  In contrast, locality cannot be exploited in the
global space: its value is in providing high bandwidth random access
to any word in a shared data structure by any processor.

% Whereas in the PGAS model coherency of global copies is implictly
% maintained across the system \todo{cite pgas?}, we forego caching of globally shared
% data.  A node can exploit temporal locality of data stored in
% the global space only by first copying it into the node's local space.
% In this way, a node can access data at random across the system while
% incurring no hardware coherence overhead: consistency is enforced by
% our mechanisms for explicit global synchronization.

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent.  Our programming model serves to
augment threaded programming languages so that they become latency
tolerant on large address spaces.  Software multithreading is used to
provide latency tolerance: a computation that is about to execute a
long latency operation such as a remote memory reference initiates the
operation and then yields to the scheduler, which quickly resumes
execution of another computation.  Concurrency is also used -- via
aggregation of requests to a common node destination -- to make more
efficient use of network bandwidth and in some cases reduce bandwidth
demands.

Synchronization on locations in the node private address space is the
same as on conventional systems.  Synchronization on global addresses
works differently: we provide atomic operations such as {\tt
  int\_fetch\_add} as well as operations using full-bits, as on the
Cray XMT.  As is true for other long latency operations,
synchronization latency is tolerated by yielding to the scheduler.
Even non-blocking global atomics yield: atomic operations executed as
x86 instructions sabbotage concurrency by forcing all outstanding
memory references to complete before new ones can be issued.  By
yielding, the core can switch to another computation while the
synchronization is accomplished outside of the pipeline.  In addition,
the implementation can exploit aggregation when concurrency is
abundant to increase the throughput of global synchronization.

Thus, the programming model provides the programmer with options: express
locality within the private space, and it will be exploited to reduce
latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
tolerating latency and reducing the cost of bandwidth.

>>>>

Our goal is to preserve the XMT programming model: large shared global
address space, explicit concurrency with a large number of threads,
and full-empty-bits synchronization.  We partition the overall address
space into a global shared address space and per node private address
spaces. Locality may be exploited directly by the programmer in the
node private address spaces, just as in a conventional cache-coherent
multiprocessor node. In contrast, locality cannot be exploited in the
global space: its value is in providing high bandwidth random access
to any word in a shared data structure by any processor.

% Whereas in the PGAS model coherency of global copies is implictly
% maintained across the system, % \todo{cite pgas?}
% we completely forego
% caching of globally shared data.

% A node can exploit temporal locality of data stored in the global
% space only by first copying it into the node's local space.  In this
% way, a node can access data at random across the system while
% incurring no hardware coherence overhead: consistency is enforced by
% our mechanisms for explicit global synchronization.

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent. The goal of our system is to support
that concurrency and consequently provide latency tolerant on a large
address space. We do so via software multithreading: computation that
is about to execute a long latency operation (e.g. a remote memory
reference) initiates the operation and then yields to the scheduler,
which quickly resumes execution of another computation.  Concurrency
is also used---via aggregation of requests to a common node
destination---to make more efficient use of network bandwidth and in
some cases reduce bandwidth demands.

Synchronization on locations in the node private address space is the
same as on conventional systems.  Synchronization on global addresses
works differently: we provide atomic operations such as {\tt
  int\_fetch\_add} as well as operations using full-empty-bits, as on
the Cray XMT.  As is true for other long latency operations,
synchronization latency is tolerated by yielding to the scheduler.
Even non-blocking global atomics yield so the core can switch to
another computation while the synchronization is accomplished outside
of the pipeline.  In addition, the implementation can exploit
aggregation when concurrency is abundant to increase the throughput of
global synchronization.
